Modeling the Evolution of Mimicry

### 2016-02-25 docs/paper/Paper.tex
Before removing all tables

%A few paragraphs are required to bring an appropriate introduction.
Mimicry is a process of deception. It is an evolutionary process with the help of which organisms survive by deceiving its predator. But this deception happens only if the environment contains similar appearing noxious organisms which the predators find unpalatable. Palatable organisms mimic the unpalatable ones through the process of evolution for survival of its species.

Artificial Intelligence is the field of computer science where scientists take concepts from psychology and implement into computer science. In contrast to that, in Artificial Life scientist take concepts from biology, specially evolutionary biology and genetics and implement into computer science.

Christopher G. Langton is considered as one of the founders of the field of Artificial Life. He invented the term while organizing the first `\textsl{International Conference on the Synthesis and Simulation of Living Systems'} at the Los Alamos National Laboratory in 1987.

In defining Artificial Life Langton states the following,

\begin{quote}
\textsl{`` `Art' + `Life' = Artificial Life: Life made by Man rather than by Nature. Our technological capabilities have brought us to the point where we are on the verge of creating `living' artifacts. The field of Artificial Life is devoted to studying the scientific, technological, artistic, philosophical, and social implications of such an accomplishment."}
\end{quote}

\gls{al} has been described by Taylor, also as a tool for biological inquiry \citep{taylor1993}. While providing a brief survey over different AL models he talks about \textsl{Wetware systems} which work at the molecular level, the \textsl{Software systems} which work at the cellular level and the \textsl{Hardware systems} which works at the organism level. \textsl{Wetware} artificial life systems are the most similar to natural life and indeed are actually derived from natural life. Most of the experiments are attempts to direct an artificial evolutionary process toward the production of \gls{rna} molecules with specific catalytic properties. The initial contribution of \textsl{software systems} in artificial life was from John von Neumann when he designed the first artificial-life model (without referring to it as such), the famous self-reproducing, computation-universal cellular automata \citep{neumann1966}. He tried to understand the fundamental properties of living systems, especially self-reproduction and the evolution of complex adaptive structures, by constructing simple formal systems that exhibit those properties. \textsl{Hardware systems} at the organism level can deal with an organism's sensory and nervous system, its body and its environment. Even though an organism's nervous system is usually considered as fantastically complex, the other interesting complex components can be its geometric, mechanical, dynamical or thermal properties.

\subsubsection{Complex Adaptive System}
\label{subsubsec:complex-adaptive-system}
\gls{cas} are special cases of complex systems. They are complex as they are diverse and made up of multiple interconnected elements and adaptive as they have the capacity to change and learn from experience. Examples of complex adaptive systems include the stock market, social insect and ant colonies, the biosphere and the ecosystem, the brain and the immune system, the cell and the developing embryo, manufacturing businesses and any human social group-based endeavor in a cultural and social system such as political parties or communities. This includes some large-scale online systems, such as collaborative tagging or social bookmarking systems.

\subsubsection{Echo}
\label{subsubsec:echo}
Echo is a class of simulation model for complex adaptive system, designed primarily for gedanken experiments rather than precise simulations. Echo provides a population of evolving, reproducing agents distributed over a geography, with different inputs of renewable resources at various sites. Each agent has simple capabilities - offense, defense, trading and mate selection - defined by a set of chromosomes. Even though these capabilities are simple and defined simply, they provide a rich set of variations illustrating the four kernel properties of complex adaptive systems described in section \ref{subsubsec:complex-adaptive-system}.

\paragraph{Criteria}
Echo has been constructed based on a set of criterion. The following is a short summary on it extracted from Holland's \textsl{Hidden Order: How Adaptation Builds Complexity} \citep{holland1996}.

\begin{itemize}
	\item Simplicity is considered as the first and foremost criterion. Instead of emulating a real system Echo is more of a thought experiment. Simplicity is attained by carefully constraining agent interaction and also giving the agents only a primitive internal model. 
	\item Actions and interaction of each agent in echo has been designed to be part of a wide range of \textsl{CAS} setting. Their mobility is also part of a wide range of geographical setting. Input to the environment which is considered as a stimuli or resource, can also be diversified.
	\item Fitness of each agent, is an evolving criterion. Instead of considering fitness as an external exogenous factor, it should be dependent on the context provided by the site and other agents. The fitness value, instead of being a constant should evolve along with the system. 
	\item \textsl{``The primitive mechanisms in Echo should have ready counter parts in all \textsl{CAS}"}. Interpretation of results from the Echo model should be direct and ready-made. As simulation is nothing but manipulation of numbers and symbols, it is always possible to make fanciful and deceiving interpretation of them, taking the opportunity to prove one's point. So by applying primitive mechanism for interpretation it is possible to avoid misinterpretation. Another advantage is with the help of interpretations, selected mechanisms can be shown to be sufficient to generate the phenomenon of interest. Even though simulation cannot establish that a given mechanism is actually present, but it can establish its sufficiency or plausibility.
	\item The architecture of the Echo model should be flexible enough so that other well-studied mathematical models of particular \textsl{CAS} can be incorporated whenever required. Among other models the significant ones include, Dawkins' \textsl{biological arms race} \citep{dawkins1990} and Brower's \textsl{survival of mimics} \citep{brower1988} from ecology; Wicksell's \textsl{Triangle} \citep{marimon1989} and \textsl{overlapping generation models} \citep{anderson1989} in economics; the \textsl{prisoner's dilemma game} \citep{axelrod1984} in political science; Holland's \textsl{two-Armed bandits} \citep{holland1975} in operational research and Perelson's \textsl{antigen-body matching} \citep{perelson1999} in immunology. 
	\item As mathematical analysis provides the surest route to arrive at a valid generalization point, most aspects of the Echo model should be amenable to mathematical analysis. 
\end{itemize}

In developing Echo using the above set of criterion, a step by step approach has been taken. Among these steps of improvement, the work by Hraber titled ``The Ecology of Echo" \citep{hraber1997} is more from an ecological perspective and has been discussed in the following.

\paragraph{Organization}
As Echo is an agent-based model, it has a collection of agents distributed in a two dimensional site which contains resources. Agents in Echo have freedom and flexibility to move around in this environment. These agents are capable of replicating themselves, while also having the capability of making complex interactions. Interactions include combat, trading and mating.

\subparagraph{Resources and Sites}
Resource in a site of Echo form its very foundation. They are represented by letters (a,b,c,d). Everything in Echo is constructed by combining these letters as strings. These resources are treated much like an `atom' and by combining them as a molecular string, agents are constructed. Geographically, Echo is a collection of interconnected sites. Each site contains a fountain which is the source of resource for the agents in Echo. The amount of resource available in each site varies in time. A site with no resource can be compared to a `desert' while another site with increased availability of letters can be a water spring or a pond.

\subparagraph{Agents}
\textsl{``Agents have a genome that is roughly analogous to a single chromosome in an haploid species"}. An agents \textsl{tag} is an external phenotype of its respective genotype (such as color). Agents have internal conditions or rules of interaction, which are genetically represented and they encode the agents internal preferences, behavioral rules and so forth. So interactions between multiple agents occur based on their conditions (rules for interactions) and tags (appearance). This results in sophisticated interaction between agents such as deception (mimicry, bluffing) and in-transitivities. \textsl{``The six (external) tag and (internal) condition genes possessed by every agent are the offense tag, defense tag, mating tag, combat condition, trade condition and mating condition"}. These genes are used to determine what sort of interaction between agents will take place and also the outcome of their interaction.

Agents have two ways of reproduction, one being self replication, while the other is by interaction between a pair of agents.

\subparagraph{Replication}
The replication process for agents is asexual reproduction. An agent needs to collect sufficient resource to start the self-replication process. After replication, the newly created agent receives a fraction of its parent's resource. Spontaneous mutation is applied to the gene for this process, where different mutation rate is applied to different parts of the gene.

\subparagraph{Interactions}
There are diverse interactions between agents consisting of combat, trading and mating. Agents need to be in the same location in site for any interaction to take place. Testing for which selection of interaction to take place, is always conducted in the order of combat, trade and mating.

\subparagraph{Combat}
The interaction of combat is analogous to real world antagonistic behavior, where one agent is killed resulting in the survivor agent to receive all its resource. Depending on each agent's offense and defense tag and the tag prefixes, the winner is decided. A \textsl{combat matrix} is also calculated with the genome of individual agents, which helps in calculating the winner.

\subparagraph{Trading}
When two agents are chosen to interact and they do not engage in combat, they are given the opportunity to trade and mate. Trading and mating are very much a mutual agreement between agents compared to combat, which is more of a destructive process. \textsl{``Agents A and B will trade if A's trading condition is a prefix of B's offense tag and vice versa"}. Trading takes place for excess resources between agents, where excess is defined as the amount which the agent possess above the amount required for a replication process, plus some reserves. The value of the reserve is controlled with the help of a simulation parameter. A deceptive process is introduced here while trading takes place, when an agent might not have excess resource but will still be engaged in trade, as it is not possible for the other agent to know the resource of its opposite interactor.

\subparagraph{Mating}
Only if two agents find each other suitable, they will engage in the mating process. Suitability for mating is decided on certain conditions based on the mating tag of individual agents. A form of two point crossover takes place to perform the mating action and produce diverse new agents.

\paragraph{Discussion of Results}
The Echo model has good resemblance with natural ecosystem. It has stable patterns of interactions among genetic variants, or trophic networks. It consists of different levels of resources along with relative species abundance. Even different cataclysmic events such as meteor, drought and so fourth are introduced. Also several isolation effects are observed where a number of sites are allowed to evolve independently. In addition to that Echo contains interesting flow of resources.

Two separate experiments are done with the above simulation to observe whether species abundance pattern can be used to find ecologically plausible behavior in Echo. To make a falsifiable comparison with biological systems, two ecological patterns are considered, each of which has similar quantitative properties in many different ecosystems that exist in nature. The objective is to compare these naturally occurring ecosystems with Echo. The ecological patterns are the Preston  curve and the species-area scaling relation. This allowed comparison of Echo diversity patterns with empirical patterns of species abundance.

Preston's canonical log normal distribution is the most widely accepted formalization of the relative commonness and rarity of species. The species-area scaling relation is also an important ecological pattern. The theory of island biography is predicted on species-area relations. In conservation biology, this relation has been used to predict the effects of different reserve size, with species diversity. Although the species-area relation can be derived from Preston's canonical log-normal distribution, a satisfactory explanation of the processes that regulate this relation has not been advanced.

Now the question is, does Echo exhibit the robust ecological phenomenon of Preston curve and species area scaling relation? For Preston's curve plot, the data show, qualitatively the pattern of genome abundances in Echo population resembles the general patterns found in biological systems, although some differences are observed. But for species area curve some discrepancy was observed between the expected result and the observed values. Several factors were raised for this discrepancy. One was the choice of the simulation parameter. Second being the definition of different species in Echo as agents having different genome is considered from different species. Another factor being the balance between rates of speciation and extinction of population.

Finally to answer the above question, it has been concluded that, Echo showed good quantitative agreement with naturally occurring species abundance distribution and species area scaling relation.

@@

\subsubsection{Definitions of Mimicry}
The best known definition of mimicry quoted to the present day, are those given by Wallace and listed in Poulton's \textit{The color of animals, their meaning and use} \citep{poulton1890colours}. The clauses are presented in the following:

\begin{itemize}
	\item \textsl{``that the imitative species occur in the same area and occupy the same station as the imitated".}
	\item \textsl{``that the imitators are always the more defenseless".}
	\item \textsl{``that the imitators are always less numerous in individuals".}
	\item \textsl{``that the imitators differ from the bulk of their allies".}
	\item \textsl{``that the imitation, however minute, is external and visible only, never extending to internal characters or to such as do not affect the external appearance."} 
\end{itemize}

According to Wickler \citep{wickler1986}, these five requirements, if fulfilled, indeed increase the probability that any given case of mimicry will be correctly diagnosed. But such supplementary clauses cannot be used as a definition of mimicry in general, though this is unfortunately often attempted.

@@

According to Huheey \citep{huheey1988} \textsl{``Mullerian mimicry is normally viewed as a system of mutual warning coloration arising from the convergence of two or more warningly colored species, though two closely related species could also be Mullerian mimics through parallel evolution" \citep{muller1879}.} In Muller's view, predation load on each species are reduced when their warning signals are shared or converged. Naive young predators create the most impact in learning predation as they start their life with no knowledge or experience over the prey species. For cases like these, all unpalatable species should benefit from mutual advertising. If mutation gives opportunity for a single unpalatable prey to resemble a warningly colored species, it is expected to be more likely to survive predation than if the change was in opposite direction. Predators surviving on warningly colored species need to generalize; meaning if similar patterns signal unpalatability, predators recognition rate becomes much higher. Thus, variation in color pattern is tolerated, and mutations leading to even poor resemblance provides some protection.

It is worth observing the fact that Mullerian mimicry is not deception, as it is to the predator's advantage to be able to recognize all deterrent and noxious species. Preys resemblance does not have to be close enough to cause misidentification on the part of the predator, but merely similar enough to remind it of a bitter past experience with similar prey. As Huheey continues \textsl{``Thus although there is selection for convergence to perhaps the most common, average pattern, or more likely the one associated with the most deterrence (both from the intensity of the traumatic stimuli and the frequency of predators receiving them), the convergence need not be rapid or precise. Indeed, the resemblance among Mullerian mimics is often not close, and the existence of very close mimics raises doubts that the mimicry is Mullerian."}

\begin{figure}[h]
	\centering
	\includegraphics[scale=0.75]{../tex/images/BatesMimButter}
	\caption[A very well-known example of mimicry, the viceroy butterfly]{A very well-known example of mimicry is the viceroy butterfly (top) which has a pattern very similar to the unpalatable monarch butterfly (bottom). It was considered for a long time that this is an ideal example of Batesian mimicry. But recently Brower discovered that the viceroy is actually just as noxious as the monarch, making this a case of Mullerian mimicry. \citep{brower1991} Image source: \href{http://en.wikipedia.org/wiki/Mullerian_mimicry}{Wikipedia}}
	\label{fig:mullerian-butterfly}
\end{figure}

@@

On this issue Turner states,

\textsl{``According to punctuated equilibrium, therefore, evolution is not more or less both gradual and continual, as has been supposed for most purposes in most post-Darwinian theories. Rather, most morphological change is held to take place within geologically short periods, separated by vastly longer periods of comparative stasis \citep{eldredge1972}. The rapid, large changes all occur during speciation, that is to say during the branching of the evolutionary tree (cladogenesis);" \citep{turner1988}}

\begin{figure}[h]
	\centering
	\includegraphics[scale=0.5]{../tex/images/Punctuated-equilibrium}
	\caption[Punctuated equilibrium vs. phyletic gradualism]{Punctuated equilibrium, bottom, consists of morphological stability and rare bursts of evolutionary change, while the top explains phyletic gradualism. Image source: \href{http://en.wikipedia.org/wiki/Punctuated_equilibrium}{Wikipedia}}
	\label{fig:punctuated-equilibrium}
\end{figure}

@@

\paragraph{Formation of Mimicry Rings}
In regards to the formation of mimicry ring Turner \citep{turner1988} has given a theory which takes into account the level of protection of different rings of butterfly species and their difference in phenotypic warning patterns. Different level of variation of these two parameter can give us very different formation of mimicry ring evolution. The cases are considered below.

\begin{itemize}
	\item If we imagine two different rings of similar warning patterns in one single habitat, similar enough that the envelop of protection afforded to one gives some protection to the more extreme variant of the other (in a simplified language, suppose that predators that have sampled species A sometimes avoid the more extreme members of species B because they mistake them as A), then these two species are subject to natural protection for mutual convergence of their patterns. Eventually they will form a rather accurate Mullerian mimicry ring.
	\item By contrast, if the two species have patterns so dissimilar that predators encountering one never imagine that it might be the other (the envelop of protection do not overlap the other species' phenotype distribution), then there is no selection whatever for convergence, and the two patterns will remain distinct indefinitely.
 \item But mutations of color patterns are occurring all the time. Suppose that species B happens to produce a mutation whose pattern falls within the envelop of protection afforded to species A. The mimicry need not be perfect, but if it is good enough, the mutation will have an advantage and will spread. In this way species B may, by a single mutation, become a Mullerian mimic of species A; in other word species B may switch from one Mullerian mimicry ring to another. Clearly, if the mimicry is not perfect, the two patterns will then gradually converge, making the resemblance ever closer. 
\end{itemize}

@@

%Entire section until past work needs to be revised when the model has been fully described.
The objective of this thesis is to design an agent based artificial life model for simulating the natural process of the evolution of mimicry.

There are mainly two species of agents. These agents have properties and behavior similar to the \textbf{model}, the \textbf{mimic} and the \textbf{predator} in the evolution of mimicry. We represent evolution of pattern for the model and the mimic with the help of \gls{ca}. Cellular Automata can be easily represented by simple rules, which can be expressed as a binary string. The predator will be equipped with a Hopfield network \citep{hopfield1982}, to be able to have pattern recognition capability. The process of evolution will of course be running at the genetic level.

@@

The latest work on modeling evolution of warning signals and mimicry with individual based simulation is done by Franks and Noble. Their initial work \citep{franks2002} seems to focus on putting some conditions of mimetic evolution in an individual based model with multiple species preyed upon by a single abstract predator, where the appearance of each prey species can evolve but their palatability is fixed.

On 2003 \citep{franks2003} another model for the origin of mimicry ring has been proposed by Franks and Noble. Accordingly theory suggests that all Mullerian mimics in an ecosystem should converge into one large ring, while this convergence will be encouraged by presence of Batesian mimics. So an evolutionary simulation to observe the above mentioned phenomenon has been presented in this piece of work.

Franks and Noble continue to test the influence on mimicry ring evolution by Batesian mimics in their work on \citep{franks2004}. Usually mathematical models of mimicry has fixed prey coloration and appearances, which enables a comparison of predation rates to demonstrate the level of protection a mimic might be afforded. In this model prey colorations are free to evolve. This phenomenon is used to examine the effect of Batesian mimicry on Mullerian mimics and mimicry rings.

\subsubsection{Models by Franks and Noble}
\label{subsubsec:models-by-frank-and-noble}
%Explain more about the first model by Franks and Noble
The \textbf{first} model by Franks and Noble \citep{franks2002} is where \textsl{``multiple species are preyed upon by a single abstract predator; the appearance of each prey species can evolve but their palatability is fixed"}. Each individual had a single gene: a value representing their external appearance or phenotype. The phenotypes are constrained to a ring of values from 1-20 (where 20 and 1 are neighbors). The distance of one phenotype from another represents their levels of similarity.

A single abstract predator was modeled with a simple reinforcement learning system. The predator's experience of each phenotype was represented by a score, which would be maintained by probability to consume the next prey species depending on similarity or difference in phenotype.

The existence of mimetic effect were measured with the initial and final distances between prey species' phenotypes. Three experiments were noted. Firstly, with one palatable and one unpalatable species. For this Batesian mimicry was evolved. For the second experiment, with two unpalatable species, Mullerian mimicry was evolved if the two prey species have some initial resemblance. Experiment 3 was carried out with two unpalatable and one palatable species, where the phenotype of the palatable species moved towards that of one of the unpalatable species, which is in other words Batesian Mimicry.

The \textbf{second} model by Franks and Noble \citep{franks2003} is based on two working hypothesis:

@@

%\subsubsection{Model Description}
\paragraph{Prey}
Similar to the first model this also contains a population of prey species each having an appearance and palatability level. Different species of prey were each assigned a fixed palatability level on a scale between zero and one (least to most palatable), where 0.5 is neutrally palatable. Palatable species have values greater than 0.5, and unpalatable species have values lower than 0.5. Each prey species have used two genes with values compositely representing their external appearance or phenotype. Both of these genes were constrained to values from 1 - 200. The Euclidean distance of one phenotype from another represented their level of similarity.

\paragraph{Predator}
Similar to Turner's stochastic model \citep{turner_et_al1984},  predators were modeled with a Monte Carlo reinforcement learning system. The predator's experience of each phenotype was represented by an attack probability, which was initialized to ambivalence at 0.5. After eating prey of a particular phenotype, the predator would make a post-attack update of the relevant probability according to the palatability of the prey consumed. The predator would use its experience of different prey appearances to help it decide on whether or not to attack at the next opportunity.

In contrary to the stochastic model \citep{turner_et_al1984},  predators would generalize on the basis of experience. A set of probability formula was used to come up with the current probability to consume a prey species based on its palatability and the previous probability of consumption based on experience using generalization rate and the Euclidean distance between the experienced phenotype and the consumed prey phenotype.

%Explain the results for the second model by Franks and Noble
\paragraph{Results}
According to the first experiment, which started with 20 unpalatable prey species, hypothesis 1 of a single large mimicry ring was not established. There was existence of multiple mimicry rings of very different frequency (population). Second experiment was carried out with some palatable species along with the unpalatable ones which was able to reach conditions of Batesian mimicry also with less mimicry rings, where hypothesis 2 borne out. It can be observed that,

\begin{quote}
\textsl{``The presence of Batesian mimics would provide positive selection pressure on mutants and would, therefore, increase the probability that they would evolve an initial resemblance to another unpalatable species. Also, Batesian pressure on mimicry rings has the potential to push one ring into the range of another, helping to bridge a large phenotypic difference between them." \citep{franks2003}}
\end{quote}

@@

\begin{table}[h]
	\centering
	\setlength\tabcolsep{2pt}
	\begin{tabular}{| p{2.2cm} | c | c | c | c | c | c | c | c |}
	  \hline
	  Current Pattern & 111 & 110 & 101 & 100 & 011 & 010 & 001 & 000 \\ \hline
	  New state of center cell & 0 & 0 & 0 & 1 & 1 & 1 & 1 & 0 \\
	  \hline
	\end{tabular}
	\caption{Cellular Automata rule}
	\label{tab:cellular-automata-rule}
\end{table}

In generating the pattern of figure \ref{fig:cellular-automata-rule-30}, the genetic representation would be the `New state of center cell'. As figure \ref{fig:cellular-automata-rule-30} is a rule 30 cellular automata, the eight bit binary representation of 30 would be the genetic representation for this pattern. Generation of automata pattern from this binary string can be considered analogous to the complex pathway between the genotype and the phenotype that exists in all living organism.

@@

With CA based pattern representation, population of prey species with a specific CA pattern can be grouped as one single species. And by restricting inter species reproduction we can control the diversity of patterns. But there is mutation applied when similar species mate with each other, so new species do born out of generations of existing species. So that is why we have two separate mutation rate while reproducing prey species. One being the ``Pattern Mutation Rate" (default values are mentioned in table \ref{tab:prey-control-parameters}) with which we control mutation of the first 8 bits of the genome while the ``Genome Mutation Rate" is used to control mutation of rest of the 9 bit genome (table \ref{tab:prey-genome}). Similar efforts of multiple mutation rate at varying location has been used in developing Echo as mentioned in section \ref{subsubsec:echo} \citep{hraber1997}.

\subsubsection{Genome}
The Genome of the prey species consists of 17 bits as presented in table \ref{tab:prey-genome}. The first eight bits represent the rule, which is used to generate Cellular Automata pattern. The next two bits are used to represent palatability of the organism explained detail in section \ref{subsubsec:genetic-palatability-representation}. The next six bits is the magnitude of the force with which mobility of the organism is calculated. The 17th bit is used to evaluate reproduction capability of the organism, depending on which the prey species is either capable or not capable to participate in reproduction.

@@

\subsubsection{Reflection of punctuated equilibrium}
\label{subsubsec:reflection-of-punctuated-equilibrium}
Punctuated equilibrium has been defined in section \ref{subsec:evolutionary-dynamics-of-mimicry} as the evolutionary process that is more inclined to cladogenesis instead of gradualism. Also Turner \citep{turner1988} emphasizes on punctuated equilibrium to describe the evolution of mimicry instead of phyletic gradualism. The design of the model under discussion also follows Turner's explanation in terms of evolving mimicry. As it can be observed, new CA patterns evolve from existing ones in prey population just by a single mutation in the pattern gene. Mimics do not follow a gradual process of evolution to look close to models but rather the change happens randomly through a single step mutation. The mutations that are favored, helps the mimics to survive while the unfavored ones fail to persist. It can be observed later in table \ref{tab:diff-in-pattern} how CA patterns of prey species can have vastly different configuration for a unit change in their representative gene.

@@

\subsubsection{Learning}
The objective of a predator's interaction with prey is always to consume it. But based on the prey's pattern and palatability, the predator will either be able to consume it or throw it back to the environment. At this event the predator needs to learn the pattern with which the prey has been represented. The pattern represents palatability of the prey species, at least to the predator. To store this new pattern into memory, each predator is equipped with a Hopfield Network Memory. Every time a new interaction is made by the predator its memory is initialized with all the existing pattern that has already been encountered and the new one. The learning procedure used for this memory is Hebbian Learning.

\paragraph{Hebbian Learning}
\textit{Hebb's postulate of learning} is the oldest and most famous of all learning rules; it is named in honor of the neuropsychologist Donald Hebb(1949). Hebb's book \textit{The Organization of Behavior} \citep{hebb1949} states the following:

\begin{quote}
\textsl{``When an axon of cell A is near enough to excite a cell B and repeatedly or persistently takes part in firing it, some growth process or metabolic changes take place in one or both cells such that A's efficiency as one of the cells firing B is increased."}
\end{quote}

@@

For Hebbian learning rule the learning signal is equal simply to the neuron's output. We have
\begin{equation}
	r \overset{\Delta}{=} f(\textbf{w}_i^t \textbf{x})
%\label{eq:}
\end{equation}
The increment \(\Delta\textbf{w}_i\) of the weight vector becomes
\begin{equation}
	\Delta\textbf{w}_i = cf(\textbf{w}_i^t \textbf{x})\textbf{x}
%\label{eq:}
\end{equation}
The single weight \( w_{ij} \) is adapted using the following increment:
\begin{equation}
	\Delta\textit{w}_{ij} = cf(\textbf{w}_i^t \textbf{x})x_j
%\label{eq:}
\end{equation}
This can be written briefly as 
\begin{equation}
	\Delta\textit{w}_{ij} = c o_i x_j, \> \text{for} \> j = 1, 2, ..., n
%\label{eq:}
\end{equation}
Here, \(\textbf{o}\) is the output vector. The learning rule requires the weight initialization at small random values around \( \textbf{w}_i = \textbf{0}\) prior to learning. The Hebbian learning rule represents a purely feed-forward, unsupervised learning. The rule implements the interpretation of the classic \textit{Hebb's postulate of learning} stated above.

The rule states that if the cross product of output and input, or correlation term \(o_ix_j\) is positive, this results in an increase of weight \(w_{ij}\); otherwise the weight decreases. It can be seen that the output is strengthened in turn for each input presented. Therefore, frequent input patterns will have most influence at the neuron's weight vector and will eventually produce the largest output.

@@

	\item \textit{Outputting.} Let \( \boldsymbol{x}_{fixed} \) denote the fixed point (stable state) computed at the end of the step 3, The resulting output vector \( \boldsymbol{y} \) of the network is
	\begin{equation}
		\boldsymbol{y} = \boldsymbol{x}_{fixed}
	\label{eq:}
	\end{equation}
	Step 1 is the storage phase, and steps 2 through 4 constitute the retrieval phase.
\end{enumerate}

@@

In this way the predator memory is limited to \textsl{``Hopfield Maximum Memory Size"}. After reaching this memory predator will not store any more new pattern but will try to associate with the existing ones it has already stored. Algorithm \ref{algo:algorithm-attack-prey} and \ref{algo:algorithm-attemptToKill} explains the process in every detail.

>>>>

%A few paragraphs are required to bring an appropriate introduction.
Mimicry is a process of deception. It is an evolutionary process with the help of which organisms survive by deceiving its predator. But this deception happens only if the environment contains similar appearing noxious organisms which the predators find unpalatable. Palatable organisms mimic the unpalatable ones through the process of evolution for survival of its species. The objective of this paper is to present an agent based artificial life model for simulating the natural process of the evolution of mimicry.

According to Langton, Artificial Life is `\textsl{Life made by Man rather than by Nature'}. Taylor also defines it as a tool for biological inquiry \citep{taylor1993}. While providing a brief survey over different AL models he talks about \textsl{Wetware systems} which work at the molecular level, the \textsl{Software systems} which work at the cellular level and the \textsl{Hardware systems} which works at the organism level. The initial contribution of \textsl{software systems} in artificial life was from John von Neumann when he designed the first artificial-life model (without referring to it as such), the famous self-reproducing, computation-universal cellular automata \citep{neumann1966}. He tried to understand the fundamental properties of living systems, especially self-reproduction and the evolution of complex adaptive structures, by constructing simple formal systems that exhibit those properties.

Being a special case of complex systems, \gls{cas} are diverse and made up of multiple interconnected elements and adaptive as they have the capacity to change and learn from experience. Echo is a class of simulation model for complex adaptive system, designed primarily for gedanken experiments rather than precise simulations. Echo provides a population of evolving, reproducing agents distributed over a geography, with different inputs of renewable resources at various sites. Each agent has simple capabilities - offense, defense, trading and mate selection - defined by a set of chromosomes. Even though these capabilities are simple and defined simply, they provide a rich set of variations illustrating the four kernel properties of complex adaptive systems described by Holland \citep{holland1996}.

In developing Echo using Holland's criterion, a step by step approach has been taken. Among these steps of improvement, the work by Hraber \citep{hraber1997} is more from an ecological perspective, and having a good resemblance with natural ecosystem. It has stable patterns of interactions among genetic variants, or trophic networks. It consists of different levels of resources along with relative species abundance. Even different cataclysmic events such as meteor, drought and so fourth are introduced. Also several isolation effects are observed where a number of sites are allowed to evolve independently. In addition to that Echo contains interesting flow of resources.

Two separate experiments were done to observe whether species abundance pattern can be used to find ecologically plausible behavior in Echo. To make a falsifiable comparison with biological systems, two ecological patterns are considered, each of which has similar quantitative properties in many different ecosystems that exist in nature. The objective is to compare these naturally occurring ecosystems with Echo. The ecological patterns are the Preston  curve and the species-area scaling relation. This allowed comparison of Echo diversity patterns with empirical patterns of species abundance. As result of the comparison Echo showed good quantitative agreement with naturally occurring species abundance distribution and species area scaling relation.

@@

@@

@@

@@

@@

The agent based artificial life model for simulating the evolution of mimicry, presented in this paper has mainly two species of agents. These agents have properties and behavior similar to the \textbf{model}, the \textbf{mimic} and the \textbf{predator} in the evolution of mimicry. We represent evolution of pattern for the model and the mimic with the help of \gls{ca}. \gls{ca} can be easily represented by simple rules, which can be expressed as a binary string. The predator will be equipped with a Hopfield network \citep{hopfield1982}, to be able to have pattern recognition capability. The process of evolution will of course be running at the genetic level.

@@

\subsubsection{Models by Franks and Noble}
\label{subsubsec:models-by-frank-and-noble}
The latest work on modeling evolution of warning signals and mimicry with individual based simulation is done by Franks and Noble. Their initial work \citep{franks2002} seems to focus on putting some conditions of mimetic evolution in an individual based model with multiple species preyed upon by a single abstract predator, where the appearance of each prey species can evolve but their palatability is fixed.

On 2003 \citep{franks2003} another model for the origin of mimicry ring has been proposed by Franks and Noble. This \textbf{second} model is based on two working hypothesis:

@@

Franks and Noble continue to test the influence on mimicry ring evolution by Batesian mimics \citep{franks2004}. Usually mathematical models of mimicry has fixed prey coloration and appearances, which enables a comparison of predation rates to demonstrate the level of protection a mimic might be afforded. In this model prey colorations are free to evolve. This phenomenon is used to examine the effect of Batesian mimicry on Mullerian mimics and mimicry rings.

@@

@@

With CA based pattern representation, population of prey species with a specific CA pattern can be grouped as one single species. And by restricting inter species reproduction we can control the diversity of patterns. But there is mutation applied when similar species mate with each other, so new species do born out of generations of existing species. So that is why we have two separate mutation rate while reproducing prey species. One being the ``Pattern Mutation Rate" (default values are mentioned in table \ref{tab:prey-control-parameters}) with which we control mutation of the first 8 bits of the genome while the ``Genome Mutation Rate" is used to control mutation of rest of the 9 bit genome (table \ref{tab:prey-genome}). Similar efforts of multiple mutation rate at varying location has been used in developing Echo \citep{hraber1997}.

\subsubsection{Genome}
The Genome of the prey species consists of 17 bits as presented in table \ref{tab:prey-genome}. The first eight bits represent the rule, which is used to generate Cellular Automata pattern. The next two bits are used to represent palatability of the organism. The next six bits is the magnitude of the force with which mobility of the organism is calculated. The 17th bit is used to evaluate reproduction capability of the organism, depending on which the prey species is either capable or not capable to participate in reproduction.

@@

\subsubsection{Reflection of punctuated equilibrium}
\label{subsubsec:reflection-of-punctuated-equilibrium}
Punctuated equilibrium is more inclined to cladogenesis instead of gradualism. Also Turner \citep{turner1988} emphasizes on punctuated equilibrium to describe the evolution of mimicry instead of phyletic gradualism. The design of the model under discussion also follows Turner's explanation in terms of evolving mimicry. As it can be observed, new CA patterns evolve from existing ones in prey population just by a single mutation in the pattern gene. Mimics do not follow a gradual process of evolution to look close to models but rather the change happens randomly through a single step mutation. The mutations that are favored, helps the mimics to survive while the unfavored ones fail to persist. It can be observed later in table \ref{tab:diff-in-pattern} how CA patterns of prey species can have vastly different configuration for a unit change in their representative gene.

@@

\subsubsection{Learning}
The objective of a predator's interaction with prey is always to consume it. But based on the prey's pattern and palatability, the predator will either be able to consume it or throw it back to the environment. At this event the predator needs to learn the pattern with which the prey has been represented. The pattern represents palatability of the prey species, at least to the predator. To store this new pattern into memory, each predator is equipped with a Hopfield Network Memory. Every time a new interaction is made by the predator its memory is initialized with all the existing pattern that has already been encountered and the new one. The learning procedure used for this memory is Hebbian Learning \citep{hebb1949}.

@@

For Hebbian learning rule the learning signal is equal simply to the neuron's output. We have
\begin{equation}
	r \overset{\Delta}{=} f(\textbf{w}_i^t \textbf{x})
%\label{eq:}
\end{equation}
The increment \(\Delta\textbf{w}_i\) of the weight vector becomes
\begin{equation}
	\Delta\textbf{w}_i = cf(\textbf{w}_i^t \textbf{x})\textbf{x}
%\label{eq:}
\end{equation}
The single weight \( w_{ij} \) is adapted using the following increment:
\begin{equation}
	\Delta\textit{w}_{ij} = cf(\textbf{w}_i^t \textbf{x})x_j
%\label{eq:}
\end{equation}
This can be written briefly as 
\begin{equation}
	\Delta\textit{w}_{ij} = c o_i x_j, \> \text{for} \> j = 1, 2, ..., n
%\label{eq:}
\end{equation}
Here, \(\textbf{o}\) is the output vector. The learning rule requires the weight initialization at small random values around \( \textbf{w}_i = \textbf{0}\) prior to learning. The Hebbian learning rule represents a purely feed-forward, unsupervised learning.

@@

	\item \textit{Output.} Let \( \boldsymbol{x}_{fixed} \) denote the fixed point (stable state) computed at the end of the step 3, The resulting output vector \( \boldsymbol{y} \) of the network is
	\begin{equation}
		\boldsymbol{y} = \boldsymbol{x}_{fixed}
	\label{eq:}
	\end{equation}
	Step 1 is the storage phase, and steps 2 through 4 constitute the retrieval phase.
\end{enumerate}

@@

In this way the predator memory is limited to \textsl{``Hopfield Maximum Memory Size"}. After reaching this memory predator will not store any more new pattern but will try to associate with the existing ones it has already stored.
